---
title: "Fars function in the Package fatalityR"
author: "Gustavo Bruges"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
 %\VignetteIndexEntry{Fars function in the Package fatality}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# fatalityR

This package provides utility functions to view and plot data from the US National Highway Traffic Safety Administration's Fatality Analysis Reporting System. For more details visit: https://www.nhtsa.gov/research-data/fatality-analysis-reporting-system-fars
- FARS is a nationwide census providing NHTSA, Congress and the American public yearly data regarding fatal injuries suffered in motor vehicle traffic crashes.

This vignette describes the use of the functions contained in the packages.
For this vignette you need to load the following packages:
```{r}
library(dplyr)
library(readr)
library(tidyr)
library(maps)
library(fatalityR) # This our package
```


## Reading Data
 This package includes 2013 raw data in the "inst/extdata" folder, you can access to it with the \code(system.file) function:
 
```{r}
system.file("externaldata", "accident_2013.csv.bz2", package = "fatalityR")
```
The FARS data is a annual compilation in CSV format. This packages provides the function fars_read for reading data and return tbl data frame. You can use the \code(fars_read) function to import a dataset, you only need to supply the filename parameter. See the \code(\link(fars_read)) function for more details. An example of the functionality:
```{r}

temp <- system.file("externaldata", "accident_2015.csv.bz2", package = "fatalityR")
dataset <- fars_read(temp)

head(dataset)


```

#Create an appropriate file name.

The `make_filename` provides an easy way to create the filename required. Example:
```{r}
year_file_name<- make_filename(2012)
year_file_name
```

```

# Manipulating data
fars_read_years  is a simple function that filter data in function of the years. A list of tbls representing the year, Month data extracted from the given set of years from the Fatality Analysis Reporting System.
```{r}
setwd(system.file("externaldata", package = "fatalityR"))
fars_read_years(2013:2015)

```

# Summarizing data
If you want to summarize  how much observations by year-month occurred with the \code{fars_summarize_years} function, you only have to supply a integer vector with the years you want to analyze and it will return a wide format tibble with months on rows, years on columns and number of observations as values.
```{r}
setwd(system.file("externaldata", package = "fatalityR"))
fars_summarize_years(2013:2015)
```

# Plot data
You can plot the map of the count of accidents in a specific state in a specific year using `fars_map_state`.

```{r}
setwd(system.file("externaldata", package = "fatalityR"))
fars_map_state(1,2015)
